---
title: "NCSU ST 503 HW 4"
subtitle: "Regression Exploration"
author: "Bruce Campbell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
header-includes:
   - \usepackage{bbm}
output: pdf_document
---

---
```{r setup, include=FALSE,echo=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = 'pdf')
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(tidy=TRUE)
knitr::opts_chunk$set(prompt=FALSE)
knitr::opts_chunk$set(fig.height=5)
knitr::opts_chunk$set(fig.width=7)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_knit$set(root.dir = ".")
library(latex2exp)   
library(pander)
library(ggplot2)
library(ggplot2)
library(GGally)
library(broom)
library(printr)
```



```{r}
library(mixtools)  #for ellipse
library(MASS)

n <- 200 # Number of random samples
set.seed(123)
# Target parameters for univariate normal distributions
rho <- 0.6
mu1 <- 1; s1 <- 2
mu2 <- 2; s2 <- 8

# Parameters for bivariate normal distribution
mu <- c(mu1,mu2) # Mean 
sigma <- matrix(c(s1^2, s1*s2*rho, s1*s2*rho, s2^2),
           2) # Covariance matrix

sigma

# Function to draw ellipse for bivariate normal data
ellipse_bvn <- function(bvn, alpha){
  Xbar <- apply(bvn,2,mean)
  S <- cov(bvn)
  ellipse(Xbar, S, alpha = alpha, col="red")
}

X <- mvrnorm(n, mu = mu, Sigma = sigma ) # from MASS package
plot(X,xlab="X1",ylab="X2",pch='*')
ellipse_bvn(X,.5)
ellipse_bvn(X,.1)
ellipse_bvn(X,.05)



XtX <-t(X) %*% X

XtX


```





















```{r}
df <-data.frame(X)
lm.fit <- lm(X1 ~ ., data=df)              
numPredictors <- ( ncol(df)-1)
hatv <- hatvalues(lm.fit)
lev.cut <- numPredictors *2 * 1/ nrow(df)
high.leverage <- df[hatv > lev.cut,]
pander(high.leverage, caption = "High Leverage Data Elements")
```

We've used the rule of thumb that points with a leverage greater than $\frac{2 p }{n}$ should be looked at.

### (d) Check for outliers. 

```{r}
studentized.residuals <- rstudent(lm.fit)
max.residual <- studentized.residuals[which.max(abs(studentized.residuals))]
range.residuals <- range(studentized.residuals)
names(range.residuals) <- c("left", "right")
pander(data.frame(range.residuals=t(range.residuals)), caption="Range of Studentized residuals")
p<-numPredictors
n<-nrow(df)
t.val.alpha <- qt(.05/(n*2),n-p-1)
pander(data.frame(t.val.alpha = t.val.alpha), caption = "Bonferroni corrected t-value")

outlier.index <- abs(studentized.residuals) > abs(t.val.alpha)

outliers <- df[outlier.index==TRUE,]

if(nrow(outliers)>=1)
{
  pander(outliers, caption = "outliers")
}

```

Here we look for studentized residuals that fall outside the interval given by the Bonferroni corrected t-values.

### (e) Check for influential points. 

We plot the Cook's distances and the residual-leverage plot with level set contours of the Cook distance.   
```{r}
plot(lm.fit,which =4)
plot(lm.fit,which = 5)
```

### (f) Check for structure in the model. 

Plot residuals versus predictors


```{r}

predictors <-names(lm.fit$coefficients)
predictors <- predictors[2:length(predictors)]

for(i in 1:length(predictors))
{
  predictor <- predictors[i]
  
  plot(df[,predictor],residuals(lm.fit),xlab=,ylab="Residuals",main = paste(predictor, " versus residuals", sep = ''))

}

```

Perform partial regression

```{r}
predictors <-names(lm.fit$coefficients)
predictors <- predictors[2:length(predictors)]

lm.formula <- formula(lm.fit)
response <- lm.formula[[2]] 

for(i in 1:length(predictors))
{
  predictor <- predictors[i]
  others <- predictors[  which(predictors != predictor) ]
  d.formula <-paste(response, " ~ ",sep='')
  m.formula <-paste(predictor, " ~ ",sep='')
  
  for(j in 1:(length(others)-1))
  { 
    d.formula <-paste(d.formula, others[j]," + ", sep='')
    m.formula <-paste(m.formula, others[j]," + ", sep='')
  }
  d.formula <-paste(d.formula, others[length(others)], sep='')
  d.formula <-formula(d.formula)

  m.formula <-paste(m.formula, others[length(others)], sep='')
  m.formula <-formula(m.formula)

  d <- residuals(lm(d.formula,df))
  
  m <- residuals(lm(m.formula,df))
  
  plot(m,d,xlab=paste(predictor, " residuals",sep=''),ylab="response residuals",main = paste("Partial regression plot for " , predictor,sep=''))

}
```



